Master Thesis

### 2013-05-06 thesis.tex
Moved acg stuff into acg.tex

\subsection{Abstract Categorial Grammars}

We present the grammatical framework in which we will develop our
system. Abstract categorial grammars are built upon two mathematical
structures, \emph{(higher-order) signatures} and \emph{lexicons}.

A \textbf{higher-order signature} is a set of elements that we call
\emph{constants}, each of which is associated with a type. Formally, it
is defined as a triple $\Sigma = \mathopen{<}A, C, \tau\mathclose{>}$,
where:
\begin{itemize}
  \item $C$ is the (finite) set of constants
  \item $A$ is a (finite) set of atomic types
  \item $\tau$ is the type associating mapping from $C$ to $\mathcal{T}(A)$,
    the set of types built over $A$
\end{itemize}

In our case, $\mathcal{T}(A)$ is the implicative fragment of linear and
intuitionistic logic with $A$ being the atomic propositions. This means
that $\mathcal{T}(A)$ contains all the $a \in A$ and all the $\alpha \limp
\beta$ and $\alpha \to \beta$ for $\alpha, \beta \in \mathcal{T}(A)$.

A signature $\Sigma = \mathopen{<}A, C, \tau\mathclose{>}$, by itself,
already lets us define an interesting set of structures, that is the set
$\Lambda(\Sigma)$ of \emph{well-typed lambda terms} built upon the
signature $\Sigma$. The set of well-typed terms and their types are
established through the judgment schemas in Figure
\ref{fig:type-judgments}.

\begin{figure}
  \begin{prooftree}
    \AxiomC{$\emptyset, \Gamma_i \vdash_\Sigma c : \tau(c)$ (cons)}
  \end{prooftree}
  \begin{prooftree}
    \AxiomC{$(x : \alpha), \Gamma_i \vdash_\Sigma x : \alpha$ (l-var)}
  \end{prooftree}
  \begin{prooftree}
    \AxiomC{$\emptyset, (\Gamma_i, x : \alpha) \vdash_\Sigma x : \alpha$ (i-var)}
  \end{prooftree}
  \begin{prooftree}
    \AxiomC{$(\Gamma_l, x : \alpha), \Gamma_i \vdash_\Sigma t : \beta$}
    \RightLabel{(l-abs)}
    \UnaryInfC{$\Gamma_l, \Gamma_i \vdash_\Sigma \lambda^{\circ} x. t : \alpha \limp \beta$}
  \end{prooftree}
  \begin{prooftree}
    \AxiomC{$\Gamma_l, (\Gamma_i, x : \alpha) \vdash_\Sigma t : \beta$}
    \RightLabel{(i-abs)}
    \UnaryInfC{$\Gamma_l, \Gamma_i \vdash_\Sigma \lambda x. t : \alpha \to \beta$}
  \end{prooftree}
  \begin{prooftree}
    \AxiomC{$\Gamma_l, \Gamma_i \vdash_\Sigma t : \alpha \limp \beta$}
    \AxiomC{$\Delta_l, \Delta_i \vdash_\Sigma u : \alpha$}
    \RightLabel{(l-app)}
    \BinaryInfC{$(\Gamma_l, \Delta_l), (\Gamma_i, \Delta_i) \vdash_\Sigma (t\ u) : \beta$}
  \end{prooftree}
  \begin{prooftree}
    \AxiomC{$\Gamma_l, \Gamma_i \vdash_\Sigma t : \alpha \to \beta$}
    \AxiomC{$\emptyset, \Delta_i \vdash_\Sigma u : \alpha$}
    \RightLabel{(i-app)}
    \BinaryInfC{$\Gamma_l, (\Gamma_i, \Delta_i) \vdash_\Sigma (t\ u) : \beta$}
  \end{prooftree}
  \caption{\label{fig:type-judgments} Type judgment schemas of the
    well-typed lambda terms $\Lambda(\Sigma)$ for a signature $\Sigma =
    \mathopen{<}A, C, \tau\mathclose{>}$}
\end{figure}

The definition of a well-typed lambda term already gives us an
interesting combinatorial structure. To make this structure even more
useful, we often focus ourselves only on terms that have a specific
\emph{distinguished type}. Using this notion of a signature of typed
constants and some distinguished type, we can describe languages of,
e.g. tree-like (and by extension string-like), lambda terms.

Let us take a look at an example from
\cite{pogodalla2007generalizing}. We will start with a very simple
signature of string expressions with the concatenation operator,
$\Sigma_{String} = \mathopen{<}A_{String}, C_{String},
\tau_{String}\mathclose{>}$. Our signature will need only one atomic
type, $A_{String} = \{\textsc{String}\}$. Our constants will be the
wordforms of our example fragment ($every$, $some$, $man$, $woman$,
$loves$), the empty string $\epsilon$ and the string concatention
operator, $+$. $\tau_{String}$ assigns the type $\textsc{String}$ to all
the wordforms and to $\epsilon$ and the type $\textsc{String} \limp
\textsc{String} \limp \textsc{String}$\footnote{As a convention,
  whenever we omit parentheses in a type, we presume the $\limp$ and
  $\to$ type constructors always bind from the right, meaning that $a
  \to b \to c$ should be read as $a \to (b \to c)$.}
to the concatenation operator $+$.

Now we can look at some well-typed terms from the string domain
(i.e. elements of $\Lambda(\Sigma_{String})$). The term $t_1 =
some\ +\ woman$\footnote{Here we introduce another notational
  convention. The unadorned way of writing this term would be $t_1 =
  (+\ some)\ woman$. Firstly, we will allow ourselves to drop the
  parentheses, meaning that any string of expressions $f\ x\ y$ should
  be read as $(f\ x)\ y$. Secondly, we will admit a specific infix
  notation for some constants which denote functions of two
  arguments. This will then let us write $x\ +\ y$ instead of
  $+\ x\ y$.} denotes the concatenation of $some$ and $woman$, which is
the phrase $some\ woman$; its type is $\textsc{String}$. A term such as
$t_2 = \lambda^{\circ} x. x\ +\ man$ denotes a function that appends the
string $man$ to its argument; its type is $\textsc{String} \limp
\textsc{String}$. If we restrict the set $\Lambda(\Sigma_{String})$ only
to terms having the type $\textsc{String}$, we get the set of
expressions which denote strings.

The idea of a signature is coupled with that of a \textbf{lexicon},
which is a mapping between two different signatures (mapping the
constants of one into well-typed terms of the other). Formally speaking,
a lexicon $\mathcal{L}$ from a signature $\Sigma_1 = \mathopen{<}A_1, C_1,
\tau_1\mathclose{>}$ (which we call the abstract signature) to a
signature $\Sigma_2 = \mathopen{<}A_2, C_2, \tau_2\mathclose{>}$ (which
we call the object signature) is a pair $\mathopen{<}F, G\mathclose{>}$
such that:
\begin{itemize}
\item $G$ is a mapping from $C_1$ to $\Lambda(\Sigma_2)$ assigning to
  every constant of the abstract signature a term in the object
  signature, which can be understood as its
  interpretation/implementation/realization.
\item $F$ is a mapping from $A_1$ to $\mathcal{T}(A_2)$ which links the
  abstract-level types with the object-level types that they realized
  in.
\item $F$ and $G$ are compatible, meaning that for any $c \in C_1$, we
  have $\vdash_{\Sigma_2} G(c) : \hat{F}(\tau_1(c))$ (we will be using
  $\hat{F}$ and $\hat{G}$ to refer to the homomorphic extensions of $F$
  and $G$ to $\mathcal{T}(A_1)$ and $\Lambda(\Sigma_1)$ respectively).
\end{itemize}

An abstract categorial grammar for us will then be just a collection of
signatures with their distinguished types and lexicons connecting these
signatures. A common pattern will have us using two signatures for the
surface forms of utterances (strings) and their logical forms (logical
propositions) and an abstract signature which is connected to both of
the object signatures via lexicons. Parsing is then just a matter of
inverting the surface lexicon to get the abstract term and then applying
to it the logical lexicon. Generation is symmetric, we simply invert the
logical lexicon and apply the surface lexicon.

\subsubsection{Example ACG}

We will look at an example of an ACG from

$\Sigma_{Synt}$ will be our abstract signature with $A_{Synt} = \{np, n,
s\}$, $C_{Synt} = \{C_{every}, C_{some}, C_{love}, C_{man}, C_{woman}\}$
and $\tau_{Synt}$ as follows:
\begin{itemize}
\item
\end{itemize}

...

\subsection{Interaction Grammars}

>>>>

\input{acg}
